### 2022-09-27 main/chapter_context.tex
improve ewk section

Gauge field theory describes this standard model with the symmetry group
$SU\!(3)_C \times SU\!(2)_L \times U\!(1)_Y$,
meaning that its Langrangian is invariant under local operations of the group
on its fields in certain representations.
Gluons correspond to the eight generators of $SU\!(3)_C$, whose subscript $_C$
indicates that they operate on colour charge triplets.
Weak and electromagnetic forces are unified by electroweak interactions of
$SU\!(2)_L$ and $U\!(1)_Y$, whose
four bosons $W^1$, $W^2$, $W^3$, and $B$ respectively would be massless
if the symmetry were unbroken,
but it is `spontaneously' broken with the
Higgs mechanism~\cite{
higgs1964broken,
englert1964broken
}%
\footnote{%
Historical accuracy prefers `Brout–Englert–Higgs mechanism', but to
properly credit associated work, Peter~Higgs has reportedly suggested
``ABEGHHKtH mechanism''~\cite{close2011infinity}.%
}

\TODO{L and Y, charge Q = T3 - Y or whatever it is}

without the Higgs mechanism, which inserts four scalar fields and
`spontaneously breaks' the gauge symmetry with a potential that moves the
ground state of the model to a gauge-asymmetric configuration.
From this new ground state, linear combinations of the electroweak fields
acquire masses while ``eating up'' three of the four scalar fields to yield
the three weak bosons and the photon:
$(W^1, W^2)$ mix into $(W^+, W^-)$, and $(W^3, B)$ mix into $(Z, \gamma)$.
The fourth and final scalar field is the Higgs boson $h$~\cite{
rubakov1999classical,
cottingham2007greenwood
}.

the weak and electromagnetic interactions
with the Higgs mechanism, which simultaneously provides the weak boson masses
and adds massive Higgs boson, whose Yukawa interactions with the fermions also
implement their masses~\cite{thomson2013modern}

\TODO{ABEGHHK'tH~\cite{close2011infinity}}

\TODO{
Glashow-Salam-Weinberg~\cite{
glashow1959renorm,
weinberg1967model,
salam1959weak
}
}

\TODO{charm november revolution~\cite{bjorken1985november}}

The fermions of the Standard Model are listed in Table~\TODO{}.
Older Standard Models included only two fermion generations and no
neutrino masses~\cite{
wells2020discovery
},\TODO{cite more}
but a modern model must include both three generations and non-zero neutrino
masses to be compatible with experiment:
three fermion generations are directly established
\TODO{kaon mixing CPV~\cite{
cronin1964evidence,
kabayasji1973cpv
}, direct observation of tau/top/bottom~\cite{
perl1977evidence,
herb1977observation,
abachi1995observation
}, Z width~\cite{
lep2006precision
}
},
as are neutrino masses from their flavour oscillations demonstrating that
they experience proper time~\cite{
kamiokande1998measurement,
superk1998evidence,
superk1999measurement,
lsnd1998evidence
}.

>>>>

As a gauge field theory, the Standard Model has the symmetry group
$SU\!(3)_C \times SU\!(2)_L \times U\!(1)$,
meaning that its langrangian is invariant under local operations of that group
on its fields in certain representations.

Gluons correspond to the eight generators of $SU\!(3)_C$, whose subscript $_C$
indicates that they operate on colour charge triplets.

Weak and electromagnetic forces are unified in electroweak interactions with
four bosons $W^1$, $W^2$, $W^3$, and $B$ from the
$SU\!(2)_L$ and $U\!(1)_Y$ symmetries respectively.
These would be massless if the symmetry were perfectly respected, but it is
`spontaneously' broken with the Higgs mechanism~\cite{
higgs1964broken,
englert1964broken
}%
\footnote{%
Historical accuracy prefers the name `Brout–Englert–Higgs mechanism', but
Peter~Higgs has reportedly suggested
``ABEGHHKtH mechanism''~\cite{close2011infinity}
to properly credit other associated work.%
}
by adding four scalar fields with a potential that moves the ground state of
to a gauge-\emph{asymmetric} configuration.
Around this new ground state, linear combinations of the electroweak fields
acquire masses while ``eating up''~\cite{rubakov1999classical} three of the
four scalar fields to yield the three weak bosons and the photon:
$W^1$ and $W^2$ mix into $W^+, W^-$, while $W^3$ and $B$ mix into $Z$ and
$\gamma$.
The fourth scalar is the Higgs boson $h$~\cite{
glashow1959renorm,
weinberg1967model,
salam1959weak,
rubakov1999classical,
cottingham2007greenwood
}.
Simultaneously, Yukawa interactions between the Higgs field and the fermions
also implement fermion masses~\cite{thomson2013modern}.
The subscript $_L$ on $SU\!(2)_L$ indicates that the weak forces interact only
with left-chiral fermions, which are arranged into doublets as shown in
Table~\TODO{ref table}.

Older Standard Models included only two fermion generations and no
neutrino masses~\cite{
wells2020discovery,
bjorken1985november
},
but a modern model must include both three generations and non-zero neutrino
masses to be compatible with experiment:
three fermion generations are directly established
\TODO{kaon mixing CPV~\cite{
cronin1964evidence,
kabayasji1973cpv
}, direct observation of tau/top/bottom~\cite{
perl1977evidence,
herb1977observation,
abachi1995observation
}, Z width~\cite{
lep2006precision
}
},
as are neutrino masses from their flavour oscillations demonstrating that
they experience proper time~\cite{
kamiokande1998measurement,
superk1998evidence,
superk1999measurement,
lsnd1998evidence
}.
